---
title: "Assignment 5"
author: "Lucian Scher"
date: "2026-05-19"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This assignment works from Lab 5. **Setup** below is copied from lab 5. **Assignment** uses a stratified training sample for decision-tree CV/tuning so `tune_grid()` runs faster.

------------------------------------------------------------------------

## Setup

```{r packages, include = FALSE}
library(tidytext)
library(tidyverse)
library(tidymodels)
library(textrecipes)
library(naivebayes)
library(discrim) 
library(glmnet)
```

This full data set includes more possible predictors than the text alone, but for this model we will only use the text variable

```{r data}

urlfile ="https://raw.githubusercontent.com/MaRo406/EDS-231-text-sentiment/main/data/climbing_reports_model_dat.csv"
incidents_df<-readr::read_csv(url(urlfile))
```

```{r examine-data}

# How many incidents are marked as deadly?

```

Now we'll split our data into training and test portions

```{r split-data}
set.seed(321)
incidents2class <- incidents_df |> 
  mutate(deadly = factor(ifelse(
    is.na(Deadly),
    "non-deadly", "deadly"
  )))

#split our data
incidents_split <- initial_split(incidents2class, prop = 0.8, strata = deadly)

incidents_train <- training(incidents_split)
incidents_test <- testing(incidents_split)

```

We use recipe() to specify the predictor and outcome variables and the data.

```{r recipe}
incidents_rec <- recipe(deadly ~ Text, data = incidents_train)
```

Next we add some familiar pre-processing steps on our Text variable: tokenize to word level, filter to the most common words, and calculate tf-idf.

```{r pre-process}
recipe <- incidents_rec %>%
  step_tokenize(Text) %>%
  step_tokenfilter(Text, max_tokens = 1000) %>%
  step_tfidf(Text)  # lab: step_tfidf(); name Text for textrecipes >= 1.1
```

Create a tidymodels workflow to combine the modeling components

```{r workflow}
incidents_wf <- workflow() %>%
  add_recipe(recipe)
```

```{r nb-spec}
nb_spec <- naive_Bayes() %>%
  set_mode("classification") %>% #set modeling context
  set_engine("naivebayes") #method for fitting model

```

Time for model evaluation. We'll do a standard resampling process to evaluate our naive Bayes model. Here we create 10-fold cross-validation sets, and use them to estimate performance. `vfold_cv` let's us do this.

```{r}
set.seed(123)
incidents_folds <- vfold_cv(incidents_train)
```

```{r nb-workflow}
nb_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(nb_spec)
```

To estimate its performance, we fit the model to each of the resampled folds, and then evaluate on the heldout part of each resampled fold.

```{r fit-resamples}

nb_rs <- tune::fit_resamples(
  nb_wf,
  incidents_folds,
  control = control_resamples(save_pred = TRUE)
)
```

Extract the relevant information using collect_metrics() and collect_predictions() and examine the performance metrics.

```{r performance}
nb_rs_metrics <- collect_metrics(nb_rs)
nb_rs_predictions <- collect_predictions(nb_rs)
nb_rs_metrics 
```

We'll use two performance metrics: accuracy and ROC AUC. Accuracy is the proportion of the data that is predicted correctly. The ROC curve plots the true positive rate against the false positive rate; AUC closer to 1 indicates a better-performing model, while AUC closer to 0.5 indicates a model that does no better than random guessing.

```{r performance-plot}
nb_rs_predictions %>%
  group_by(id) %>%
  roc_curve(truth = deadly, .pred_deadly) %>%
  autoplot() +
  labs(
    color = "Resamples",
    title = "ROC curve for Climbing Incident Reports"
  )
```

Another model method involves the confusion matrix. A confusion matrix tabulates a model's false positives and false negatives for each class.

```{r confusion-matrix}
conf_mat_resampled(nb_rs, tidy = FALSE) %>% #compute matrix for each fold then average
  autoplot(type = "heatmap")
```

Hmm, let's try a different model. 

Lasso classification uses regularization on regression to produce a simpler, more generalizable model. It learns how much of a penalty to apply to features, reducing the high-dimensional space of tokens — sometimes shrinking coefficients all the way to 0, effectively removing those features from the model entirely.

The penalty is a hyperparameter we need to choose. Rather than setting it manually, we estimate its best value by fitting models across 30 candidate penalty values on our resampled data and comparing performance.

```{r penalty-tuning-specification}
tune_spec <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_mode("classification") %>%
  set_engine("glmnet")

tune_spec

```

```{r lamba}
# mixture = 1 is fixed in tune_spec (lasso); grid only varies penalty
lambda_grid <- grid_regular(penalty(), levels = 30)

lambda_grid

```

Here we use grid_regular() to create 30 possible values for the regularization penalty. Then tune_grid() fits a model at each of those values.

```{r tune}
tune_wf <- workflow() %>% 
  add_recipe(recipe) %>% 
  add_model(tune_spec)

set.seed(2023)
tune_rs <- tune_grid(
  tune_wf,
  incidents_folds,
  grid = lambda_grid,
  control = control_resamples(save_pred = T)
  
)

```

```{r plot_metrics}
collect_metrics(tune_rs)
autoplot(tune_rs) + 
  labs(
    title = "Lasso performance across regularization penalties"
  )
```

```{r penalty-show-best}

tune_rs %>%
  show_best(metric = "roc_auc")

tune_rs %>%
  show_best(metric = "accuracy")

chosen_acc <- tune_rs %>% 
  select_by_one_std_err(metric = "accuracy", -penalty)
```

Next, let's finalize our workflow with this particular regularization penalty. This is the regularization penalty that our tuning results indicate give us the best model.

```{r final-model}
final_lasso <- tune_wf %>%
  finalize_workflow(chosen_acc)

final_lasso

```

The penalty argument value now reflects our tuning result. Now we fit to our training data.

```{r}
fitted_lasso <- fit(final_lasso, incidents_train)
```

First let's look at the words associated with an accident being non-deadly.

```{r words-non-deadly}
# fitted_lasso 

```

And now the words that are most associated with a deadly incident.

```{r words-deadly}
# fitted_lasso 
```

Finally, let's fit to the test data and see how we did.

```{r}
last_fit(final_lasso, incidents_split) %>%
  collect_metrics()
```

------------------------------------------------------------------------

## Assignment

**Third classifier:** Decision tree using `rpart`.

CV and hyperparameter tuning below use a **25% stratified sample** of `incidents_train` and **5-fold CV** so `tune_grid()` finishes in reasonable time. Task 3–4 `last_fit()` uses the **full** `incidents_split`.

```{r assignment-packages}
library(rpart)
```

```{r assignment-sample}
# Stratified sample for faster decision-tree CV/tuning only (full data used in last_fit)
set.seed(99)
incidents_train_sample <- incidents_train |>
  group_by(deadly) |>
  slice_sample(prop = 0.25) |>
  ungroup()

set.seed(123)
incidents_folds_sample <- vfold_cv(incidents_train_sample, v = 5)
```

### Task 1 — Cross-validated performance (default decision tree)

Claude helped me set up the default decision tree specification and workflow below. I asked it to use the `rpart` engine because it is a common implementation of decision trees in R.

```{r dt-spec-default}
# build default decision tree  
dt_spec_default <- decision_tree() %>%
  set_mode("classification") %>%
  set_engine("rpart")
```

```{r dt-workflow-default}
# Same as lab workflow but with the default decision tree spec
dt_wf_default <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(dt_spec_default)
```

```{r dt-fit-resamples}
# 5-fold CV on 25% sample
dt_rs_default <- fit_resamples(
  dt_wf_default,
  incidents_folds_sample,
  control = control_resamples(save_pred = TRUE)
)
```

```{r dt-cv-metrics}
# Mean accuracy and ROC AUC across resamples
dt_cv_metrics <- collect_metrics(dt_rs_default)

print(dt_cv_metrics)
```

**Task 1 — Summary**

Our cross-validated accuracy is 0.881 and roc_auc is 0.837. Which is better than the naive Bayes CV performance. But our decision tree CV is on a smaller sample and fewer folds, so this is just a rough baseline comparison.

### Task 2 — Hyperparameter tuning

```{r dt-spec-tune}
# Mark cost_complexity and tree_depth as tunable
dt_spec_tune <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune()
) %>%
  set_mode("classification") %>%
  set_engine("rpart")
```

```{r dt-grid}
# Build 5 x 5 grid of cost_complexity and tree_depth values to try
dt_param_grid <- grid_regular(
  cost_complexity(),
  tree_depth(),
  levels = 5
)
```

Claude suggested tuning these two hyperparameters because they control the complexity of the decision tree. I asked it to create a grid of 5 x 5 = 25 combinations of these hyperparameters to try during tuning.

```{r dt-tune}
# Fit each grid point with 5-fold CV on the sample
dt_wf_tune <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(dt_spec_tune)

set.seed(456)
dt_tune_rs <- tune_grid(
  dt_wf_tune,
  incidents_folds_sample,
  grid = dt_param_grid,
  control = control_resamples(save_pred = TRUE)
)
```

```{r dt-tune-plot}
# Visualize how accuracy / ROC AUC change across the grid
autoplot(dt_tune_rs) +
  labs(title = "Decision tree CV performance across hyperparameters (sample)")
```
Claude helped me figure out the best hyperparameter values based on the tuning results. I asked it to use the "one standard error rule" to select a simpler tree that is within 1 std error of the best ROC AUC, to help reduce overfitting.
```{r dt-show-best}
# Top-performing penalty/depth combinations on the sample
dt_tune_rs %>% show_best(metric = "roc_auc")
dt_tune_rs %>% show_best(metric = "accuracy")
```

```{r dt-choose-params}
# Choose the simplest tree within 1 std error of the best ROC AUC 
chosen_dt <- dt_tune_rs %>%
  select_by_one_std_err(metric = "roc_auc", cost_complexity, tree_depth)

final_dt <- dt_wf_tune %>%
  finalize_workflow(chosen_dt)

final_dt
```

**Task 2 — Hyperparameter explanation**

We tune cost_complexity and tree_depth. `Cost_complexity` penalizes higher values resulting in simpler trees that are less likely to memorize rare word patterns. `Tree_depth` controls how deep the tree can grow, we tune to find a balance between capturing word interactions and overfitting sparse text features. 


### Task 3 — Test-set evaluation and model comparison

```{r dt-test-metrics}
# Refit tuned tree on all training rows
dt_test_metrics <- last_fit(final_dt, incidents_split) %>%
  collect_metrics() %>%
  mutate(model = "Decision tree")

dt_test_metrics
```

```{r nb-test-metrics}
# Same last_fit pattern for Naive Bayes
nb_test_metrics <- last_fit(nb_wf, incidents_split) %>%
  collect_metrics() %>%
  mutate(model = "Naive Bayes")

nb_test_metrics
```

```{r lasso-test-metrics}
# Same last_fit pattern for tuned Lasso
lasso_test_metrics <- last_fit(final_lasso, incidents_split) %>%
  collect_metrics() %>%
  mutate(model = "Lasso")

lasso_test_metrics
```

```{r model-comparison}
# Side-by-side test metrics for all three classifiers
model_comparison <- bind_rows(
  nb_test_metrics,
  lasso_test_metrics,
  dt_test_metrics
) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

model_comparison
```

**Task 3 — Which model performs best?**
Compared to the Naive Bayes and decision tree, the Lasso has the best test performance on both accuracy and ROC AUC. The decision tree is in the middle, and Naive Bayes performs worst on the test set. I think this is because the Lasso's regularization reduces overfitting while still capturing the most important signals. As opposed to Naive Bayes which assumes feature independence and is less flexible, and the decision tree which may not capture complex interactions as well as a linear model with regularization.


### Task 4 — Confusion matrix and class imbalance

```{r dt-confusion-matrix}
# Hold-out test predictions from the tuned decision tree
dt_last_fit <- last_fit(final_dt, incidents_split)

dt_test_preds <- dt_last_fit %>%
  collect_predictions()

conf_mat(dt_test_preds, truth = deadly, estimate = .pred_class)
```

```{r dt-confusion-plot}
# Plot confusion matrix heatmap
conf_mat(dt_test_preds, truth = deadly, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "Decision tree — test set confusion matrix")
```

**Task 4a — What does the confusion matrix show?**

Our matrix reveals that the decision tree model does reasonably on the minority class but still misses about 29% of actual deadly cases. The imbalance means that the model is much better at recognizing non-deadly language than deadly language. Which could be a result of the language used in our data, that might censor or under-describe deadly incidents, making them harder to learn from. 

- **True positives:** 80 deadly incidents correctly flagged as deadly.
- **False negatives:** 33 deadly incidents predicted as non-deadly.
- **False positives:** 22 non-deadly reports flagged as deadly.
- **True negatives:** 420 non-deadly correctly classified—the majority class dominates cell counts.

The majority class (non-deadly) dominates the data, so the model is biased towards predicting that class correctly at the expense of missing deadly cases. Which is why I don't think overall accuracy is enough to evaluate model performance in this context. Instead I would prioritize recall on the deadly class, for minimizing false negatives, even if it means more false positives, because missing a deadly report is more costly than flagging an extra non-deadly one for review because we are dealing with real human lives.